Notes about Spin and TR Symmetry

### 2016-10-08 Notes about TR symmetry/Notes about TR symmetry.tex
Small fixes
Daily upate: correct error in homework. New notes about TR symmetry - kramer's theorem.

\subsection{Property of a symmetry operator}
\label{sec:Property-of-symmetry-operator}
Consider a symmetry operation acting on states such that:
\begin{align}
    \ket{\alpha}\to \ket{\widetilde{\alpha}}\,\,
    \ket{\beta}\to \ket{\widetilde{\beta}}
\end{align}
We require that the probability, the most important thing in quantum
mechanics, is unchaged:
\begin{align}
    |\braket{\widetilde{\beta}|\widetilde{\alpha}}| =
        |\braket{\beta|\alpha}|
\end{align}
Wigner has proved that\footnote{
Though it was only mentioned in page 289 of \cite{sakurai}.},
all such symmetry transformations can be characterized either by an unitary
or an antiunitary operator. Also, any antiunitary operator $\theta$ can be
expressed as
\begin{align}
    \theta = U K
\end{align}
where $U$ is unitary, and $K$ is complex conjugation. Three facts about
$K$:
\begin{itemize}
    \item \textit{"acts only on any coefficient that multiplies a ket
        (and stands on the right of K)."} 
    \item $K$ also does not act on the base ket, i.e. $K\ket{e}=\ket{e}$.
    \item $K$'s action depends heavily the base ket. This is obvious from
        the above two points. Of course, this is not important when every
        physical state is only projective. However, I still doubt whether
        this will call additional problem.
        \footnote{Consider $K\ket{e}=\ket{e}$, but when
        $\ket{e}=i\ket{e'}$, then $K\ket{e}=-i\ket{e'}=-\ket{e}$.}
\end{itemize}

@@

    \subsection{From classical notion to property of TR symmetry}
    \label{sec:From-classical-notion-to-property-of-TR-symmetry}
    To get the time reversal operator, we can first look at the following
    equality (let $\Theta$ dentoes the time reversal operator):
    \begin{align}
        \left(1-\frac{iH}{\hbar}\delta t\right) \Theta \ket{\alpha}
        = \Theta\left(1-\frac{iH}{\hbar}(-\delta t)\right)\ket{\alpha}
        \label{eq:TRoper.fromClas2Quam.phyPic}
    \end{align}
    This follows from the following notion: a time reversed state is
    equivalent to the original state at an earlier time. Recall that
    $\left(1-\frac{iH}{\hbar}\delta t\right)$ produces an infinitesimal
    time translation.

@@

    It is also interesting to note that in general:
    \begin{align}
        \Theta^2 &= -1 \text{ if $j$ is a half-integer} \\
        \Theta^2 &= 1 \text{ if $j$ is an integer}
    \end{align}
    However, this can only be proved when $j=\frac{1}{2}$, in which case
    the proof is also straightforward if we use the formula mentioned in
    section \ref{sec:About-exp(-i*pi-S_y/hbar)}.

    However, if we imagine a spin $j$ system as composed all just of
    electrons of spin $\frac{1}{2}$, then we have a strong argument that
    the above formula is true.
    \subsection{Kramer's theorem and the lifting of degenracy}

\section{Anchor}
\label{sec:Anchor}

>>>>

\subsection{Properties of a symmetry operator}
\label{sec:Property-of-symmetry-operator}
Consider a symmetry operation acting on states such that:
\begin{align}
    \ket{\alpha}\to \ket{\widetilde{\alpha}}\,\,
    \ket{\beta}\to \ket{\widetilde{\beta}}
\end{align}
We require that the probability, the most important thing in quantum
mechanics, is unchaged:
\begin{align}
    |\braket{\widetilde{\beta}|\widetilde{\alpha}}| =
        |\braket{\beta|\alpha}|
\end{align}
Wigner has proved that\footnote{
Though it was only mentioned in page 289 of \cite{sakurai}.},
all such symmetry transformations can be characterized either by an unitary
or an antiunitary operator. Also, any antiunitary operator $\theta$ can be
expressed as
\begin{align}
    \theta = U K
\end{align}
where $U$ is unitary, and $K$ is complex conjugation. Three facts about
$K$:
\begin{itemize}
    \item \textit{"acts only on any coefficient that multiplies a ket
        (and stands on the right of K)."} 
    \item $K$ also does not act on the base ket, i.e. $K\ket{e}=\ket{e}$.
    \item $K$'s action depends heavily the base ket. This is obvious from
        the above two points. Of course, this is not important when every
        physical state is only projective. However, I still doubt whether
        this will call additional problem.
        \footnote{Consider $K\ket{e}=\ket{e}$, but when
        $\ket{e}=i\ket{e'}$, then $K\ket{e}=-i\ket{e'}=-\ket{e}$.}
\end{itemize}

@@

    \subsection{From classical notion to properties of TR symmetry}
    \label{sec:From-classical-notion-to-property-of-TR-symmetry}
    To get the time reversal operator, we can first look at the following
    equality (let $\Theta$ dentoes the time reversal operator):
    \begin{align}
        \left(1-\frac{iH}{\hbar}\delta t\right) \Theta \ket{\alpha}
        = \Theta\left(1-\frac{iH}{\hbar}(-\delta t)\right)\ket{\alpha}
        \label{eq:TRoper.fromClas2Quam.phyPic}
    \end{align}
    This follows from the following notion: a time reversed state is
    equivalent to the original state at an earlier time. Recall that
    $\left(1-\frac{iH}{\hbar}\delta t\right)$ produces an infinitesimal
    time translation.

@@

    It is also interesting to note that in general:
    \begin{align}
        \label{eq:TRoper.TRoper.square_minus}
        \Theta^2 &= -1 \text{ if $j$ is a half-integer} \\
        \Theta^2 &= 1 \text{ if $j$ is an integer}
        \label{eq:TRoper.TRoper.square_plus}
    \end{align}
    However, this can only be proved when $j=\frac{1}{2}$, in which case
    the proof is also straightforward if we use the formula mentioned in
    section \ref{sec:About-exp(-i*pi-S_y/hbar)}.

    However, if we imagine a spin $j$ system as composed all just of
    electrons of spin $\frac{1}{2}$, then we have a strong argument that
    the above formula is true.
    \subsection{Kramer's theorem and the lifting of degenracy}
    When one is concerned with the energy spectrum, one encounters the
    Kramer's theorem, which says:
    \begin{thm}[Kramer's theorem]
        For a system with halt-integer $j$ spin, the degree of degeneracy
        is at least two.
    \end{thm}
    \begin{proof}
        Let $\ket{n}$ be an energy eigenstate. If the system is time
        reversal symmetric, i.e. $[H,\Theta]=0$, then $\Theta \ket{n}$ is
        another energy eigenstate with the same energy. Supposing there is
        no degeneracy, then we have
        $$ \ket{n} = e^{i\eta}\Theta\ket{n}$$
        for some $\eta\in \mathbb{R}$. Apply $\Theta$ on both sides, then
        we have
        \begin{align*}
            e^{-i\eta} \Theta^2 \ket{n} &= \pm e^{-i\eta}\ket{n} \\
            &= \Theta \ket{n} = e^{i\eta} \ket{n}
        \end{align*}
        Clearly this is a contracdiction unless $\Theta^2=1$, which is
        true only for interger spin system by equation
        \ref{eq:TRoper.TRoper.square_plus} and
        \ref{eq:TRoper.TRoper.square_minus}.
    \end{proof}
    Another way to show this is presented in page 37 of
    \cite{bernevig}. That method is too troublesome and basis dependent 
    \footnote{Borrowing the jargon of mathematicians, not canonical,
        geometric, or intrinsic.}
    to be shown here.
    But it shows an interesting fact that $\Theta^2=-1$, actually
    implies that $U$, its unitary component mentioned in section
    \ref{sec:Property-of-symmetry-operator}, is antisymmetric. This is
    obvious when one juxtapose the following two facts:
    $$ U (U^*)^T = \mathbbm{1}$$
    $$ \Theta^2 = UKUK = U U^* = -\mathbbm{1} $$

    By Kramer's theorem, one also says that for system with half-integer
    spin, one \textit{lift the degenracy by breaking the time reversal
    symmetry of the system.}

    An additional point by Bernevig is about the transition probability
    $\bra{\Theta \psi}H\ket{\psi}$ from a state to its time reversed
    state.

    \begin{align*}
        & \braket{\Theta\psi|H|\psi} = (U_{mp}K\psi_p)^*H_{mn}\psi_n 
        =U_{mp}^* \psi_p H_{mn} \psi_n = (U^\dagger)_pm\psi_p H_{mn}\psi_n
        = (U^\dagger)_{pm}\psi_p (THT^{-1})_{mn} \psi_n
        \\
        &= (U^\dagger)_{pm}\psi_p \left(U_{mr}KH_{rq}(-U_{qn}K)\right)\psi_n\\
        &= - (U^\dagger)_{pm}\psi_p \left(U_{mr}H^*_{rq}U^*_{qn}\right)\psi_n\\
        &= - \psi_p \left((U^\dagger)_{pm}
            U_{mr}\right)H^*_{rq}U^*_{qn}\psi_n\\
        &= - \psi_p H^*_{pq}U^*_{qn}\psi_n\\
            \overset{\text{by ...}}{=}& - \psi_p U^*_{pq}H_{qn}\psi_n \\
        &= U^*_{qp}\psi_p H_{qn} \psi_n
    \end{align*}
    Either there is a mistake in Bernevig's derivation or I am wrong...
\section{Anchor}
\label{sec:Anchor}
